---
title: "Pythia_analysis"
date: "2025-01-08"
output: 
  html_document:
    code_folding: hide
    theme: 'lumen'
    df_print: 'paged'
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggbeeswarm)
library(pwalign)
library(effsize)
library(Peptides)
library(kldtools)
library(ggpubr)
library(GOSemSim)
library(kableExtra)

## Load general-use datasets
### tissue specific genes. (check Bgee processing rmd to know how to get these)
orth_list<-readRDS("./RDS/large_fish_tree_all_orders.rds")
orth_list<-orth_list %>% mutate(orths =  str_remove(orth_list$orths,"..$"))
non_spuroius_conv<-read_csv("Datasets/DatasetS0_csv_files/Non_spurious_conv.csv")
combined_arity <-readRDS("./RDS/combined_arity.rds")
tau_dat_all<-readRDS("./RDS/tau_dat_all.rds")
danio_conv<-read.csv("./Datasets/DatasetS0_csv_files/Danio_rerio_convergent_genes.csv")

```

## Summary of convergent amino acid substitutions {.tabset}
In this section I functionally characterise the different convergent amino acid substitutions we observed.  
Checking the BLOSUM62 scores between the convergent substitution and other non-convergent substitutions will provide insight into the potential functional nature of the convergent sites. If the convergent sites are associated with functioning of the protein then we would expect substitutions on these sites to have lower BLOSUM62 score as compared to the non convergent ones. On an average we do observe that convergent sites have a less score than other non-convergent substitutions occurring in the same protein. When looking at the changes in specific physiochemical properties of amino acids, we found that most of the changes in amino acid properties occur during the convergent substitutions. 
To determine the changes in amino acid properties I compared the values for properties of the ancestral amino acid and compared it with the values of extant amino acids caused by the convergent substitutions. We compared 17 different amino acid properties and found that 95% of the convergent orthogroups (85 out of 89) had significant changes (two-sided Wilcoxon rank sum test FDR < 0.05) in at least one of the 17 physicochemical properties analysed. I took all the significant changes between ancestral and extant amino acid and looked at the number caused by convergent and non-convergent substitutions. Most of the changes resulted from convergent substitutions.  
```{r, eval=F}
#Get amino acid properties
data("AAdata")
rbind(
AAdata$zScales$Z1 %>% as_tibble() %>% mutate(property="Lipophilicity") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$zScales$Z2 %>% as_tibble() %>% mutate(property="Steric") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$zScales$Z3 %>% as_tibble() %>% mutate(property="Polarity") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF1 %>% as_tibble() %>% mutate(property="Helix_bend_preference") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF2 %>% as_tibble() %>% mutate(property="Side-chain_size") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF3 %>% as_tibble() %>% mutate(property="Extended_structure_preference") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF4 %>% as_tibble() %>% mutate(property="Hydrophobicity") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF5 %>% as_tibble() %>% mutate(property="Double-bend_preference") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF6 %>% as_tibble() %>% mutate(property="Partial_specific_volume") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF7 %>% as_tibble() %>% mutate(property="Flat_extended_preference") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF8 %>% as_tibble() %>% mutate(property="Occurence_in_alpha_region") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$kideraFactors$KF10 %>% as_tibble() %>% mutate(property="Surr_hydrophobicity") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$FASGAI$F2 %>% as_tibble() %>% mutate(property="Alpha_turn_prop") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$FASGAI$F3 %>% as_tibble() %>% mutate(property="Bulk_prop") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$FASGAI$F4 %>% as_tibble() %>% mutate(property="Compositional_characters") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$FASGAI$F5 %>% as_tibble() %>% mutate(property="Local_flex") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V")),
AAdata$FASGAI$F6 %>% as_tibble() %>% mutate(property="Electronic_prop") %>% 
  mutate(amino_acid= c("A","R","N","D","C","Q","E","G","H","I","L","K","M","F","P","S","T","W","Y","V"))
)->aa_properties

#Distribution of aa physiochemical properties
aa_properties %>% ggplot(aes(x=value))+
  facet_grid(cols=vars(property))+
  geom_density(fill="green4",alpha=0.6)+
  theme_bw()

sub_Matix_score<-function(id){
#Prepare csusbt site output
files<-list.files(path = paste0("./Datasets/DatasetS6_Non_spurious_convergence"), pattern = paste0(id,".tsv"),
                    full.names = TRUE, recursive = TRUE)
#read in file
dat<-read_tsv(files[1]) %>% dplyr::select(OCNany2spe, OCNany2any, contains(c("aa", "codon_site_pdb"))) %>% 
  filter(OCNany2any >0) %>% arrange(desc(OCNany2spe))
#get branches info
branches<-data.frame(branches=
                       dat %>% dplyr::select(matches("^aa(?!.*(anc|hydrophobicity)).*$", perl = T)) %>% 
                       dplyr::select(-last_col()) %>% colnames())
branches<-branches %>% filter(str_count(branches, "_") == 1)

#Special selecting function
select_grouped_columns <- function(data, prefix) {
  data %>%
    dplyr::select(OCNany2spe, OCNany2any, matches(paste0("^", prefix, "(?:_anc)?(?:_hydrophobicity)?$")))
}

make_aa_sub_dat<-function(branches){
  select_grouped_columns(dat, branches) %>% mutate(type_of_sub=case_when(OCNany2spe >= 0.5 ~ "Convergent", .default = "Non-convergent")) %>%
    dplyr::select(3,4,type_of_sub) %>% dplyr::rename("extant"=1, "anc"=2) %>% mutate(branch.num=branches)->aa_subs_data
  return(aa_subs_data)
}
#data frame with ancestral to extant aa substitution
purrr::map(branches$branches, make_aa_sub_dat) %>% list_rbind() %>% drop_na()->aa_subs_data

#Get the Blosum matrix
data(BLOSUM62)
aa_subs_data %>% rowwise() %>% mutate(sub_score = BLOSUM62[anc, extant]) %>% ungroup() %>% mutate(pdb_id=id) ->score

#Make data frame with BLOSUM scores and aa properties
rbind(
left_join(
score, aa_properties %>% filter(property=="Lipophilicity") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Lipophilicity") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
  score, aa_properties %>% filter(property=="Steric") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Steric") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
  score, aa_properties %>% filter(property=="Polarity") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Polarity") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
  score, aa_properties %>% filter(property=="Helix_bend_preference") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Helix_bend_preference") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
  score, aa_properties %>% filter(property=="Side-chain_size") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Side-chain_size") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
  score, aa_properties %>% filter(property=="Extended_structure_preference") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Extended_structure_preference") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
  score, aa_properties %>% filter(property=="Hydrophobicity") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Hydrophobicity") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
left_join(
score, aa_properties %>% filter(property=="Double-bend_preference") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Double-bend_preference") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Partial_specific_volume") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Partial_specific_volume") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Flat_extended_preference") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Flat_extended_preference") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Occurence_in_alpha_region") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Occurence_in_alpha_region") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Surr_hydrophobicity") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Surr_hydrophobicity") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Alpha_turn_prop") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Alpha_turn_prop") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Bulk_prop") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Bulk_prop") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Compositional_characters") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Compositional_characters") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Local_flex") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Local_flex") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y)),
  left_join(
score, aa_properties %>% filter(property=="Electronic_prop") %>% dplyr::rename("extant"="amino_acid"), by ="extant") %>% 
  left_join(aa_properties %>% filter(property=="Electronic_prop") %>% dplyr::rename("anc"="amino_acid"), by ="anc") %>% 
  dplyr::rename("extant.value"="value.x", "anc.value"="value.y", "property"="property.x") %>% dplyr::select(-c(property.y))
)->scores_and_properties

#Data frame with adjusted Wilcox test pvalues
#Convergent
  scores_and_properties %>% filter(type_of_sub=="Convergent") %>%
    summarise(wil_pval=p.adjust(wilcox.test(x= extant.value,y=anc.value,exact=T)$p.value, method = "fdr"), 
              .by = c(property)) %>% 
    mutate(pdb_id=id)->peptides_wil_out_conv

#Non-convergent
  scores_and_properties %>% filter(type_of_sub=="Non-convergent") %>%
    summarise(wil_pval=p.adjust(wilcox.test(x= extant.value,y=anc.value,exact=T)$p.value, method = "fdr"), 
              .by = c(property)) %>% 
    mutate(pdb_id=id)->peptides_wil_out_nc
  
  #Plot section
  make_boxplot_list<-function(property_type){
    scores_and_properties  %>% filter(type_of_sub=="Convergent", property==property_type) %>%
      dplyr::select(type_of_sub,branch.num, extant.value,anc.value,property) %>% 
      pivot_longer(cols = c(extant.value,anc.value)) %>% 
      ggplot(aes(y=value, x=name, fill=name))+
      facet_grid(rows=vars(branch.num))+
      geom_boxplot(outliers = F)+
      geom_quasirandom(shape=21, size=2.5,color="black", alpha=0.7)+
      scale_fill_manual(values = c("#1AAEC0","#EA5B15"))+
      xlab("Sequence type")+
      labs(subtitle = paste0(property_type,"\nWilcoxon p-value: ",
                             peptides_wil_out_conv %>% filter(property==property_type) %>% 
                               pull(wil_pval) %>% signif(digits = 4)))+
      theme_bw()+theme(axis.text.x = element_blank(),
                       axis.ticks.x = element_blank())->p
    return(p)
    }

map(as.list(peptides_wil_out_conv %>% distinct(property) %>% pull()), 
    make_boxplot_list)->bp_list

library(patchwork)
wrap_plots(bp_list,guides = "collect",widths = 2) %>% ggsave(filename = paste0(id,".pdf"), 
                                                             path = "./Figures/Properties_will_test/",
                                                             device = "pdf",
                                                            width = 10, height = 15, units = "in", scale = 1.5)


#return(score)
#return(combined_plots)
#return(scores_and_properties)

return(rbind(peptides_wil_out_conv %>% mutate(state="Convergent"),
             peptides_wil_out_nc %>% mutate(state="Non-convergent")))
}
#Use this to skip any error
poss_sub_Matix_score = possibly(.f = sub_Matix_score)

#For the BLOSUM matrix use return(score)
purrr::map(pdb_ids, poss_sub_Matix_score) %>% list_rbind()->score_table
#score_table %>% saveRDS("./RDS/sub_Matix_score_output_for_blosum.rds")

#Make the table
purrr::map(pdb_ids, poss_sub_Matix_score) %>% list_rbind() ->Properties_pval_output_table
#Properties_pval_output_table %>% saveRDS("./RDS/Properties_pval_output_will_test.rds")
#For direction change use return(scores_and_properties)
purrr::map(pdb_ids, poss_sub_Matix_score) %>% list_rbind()->scores_and_properties
```

### BLOSUM62 score
```{r,fig.dim=c(3,4)}
#Load table
score_table<-readRDS("./RDS/sub_Matix_score_output.rds")
#Get p-values
signif((wilcox.test(x=score_table %>% group_by(pdb_id,type_of_sub) %>% summarise(mean_score=mean(sub_score,na.rm =T), .groups = "drop") %>%
                      filter(type_of_sub=="Convergent") %>% pull(mean_score),
                    y=score_table %>% group_by(pdb_id,type_of_sub) %>% summarise(mean_score=mean(sub_score,na.rm =T), .groups = "drop") %>%
                      filter(type_of_sub=="Non-convergent") %>% pull(mean_score)))$p.value, digits = 1)->p
#Make plot
score_table %>% group_by(pdb_id,type_of_sub) %>% summarise(mean_score=mean(sub_score,na.rm =T), .groups = "drop") %>% 
  ggplot(aes(x=type_of_sub, y=mean_score, colour = type_of_sub))+
  geom_boxplot()+
  geom_quasirandom()+
  xlab("Type of substitution")+
  ylab("Mean BLOSUM62 score")+
  theme_bw()+
  theme(legend.position = "none")+
  scale_color_manual(values = c("#EA5B15","#1AAEC0"))+
  labs(subtitle = paste0("Two-sided Wilcoxon p-value: ",p))
  
```


### Distribution of convergent substitutions on protein structure
Here I check whether there is any relationship between the occurrence of convergence and site features on the protein. Using a chi-sq test of independence I found that in our data, there is no relationship between protein site feature and convergence. This tells me that there isn't a general trend for convergence to be enriched on surface residues, rather it seems specific to each protein.
```{r}
prot<-read_csv("./Datasets/DatasetS0_csv_files/protein_sites/Session_pymol/combined_prots.csv") %>% filter(Localisation != "Localisation") %>% filter(Number_aa != "Number_aa") %>% filter(Name_aa != "Name_aa") %>% filter(near_lig != "near_lig") %>% filter(Secondary_Structure != "Secondary_Structure") %>% filter(convergence != "convergence") %>% filter(Localisation != "no_build") %>% filter(Secondary_Structure != "no_build")

chisq.test(table(prot$Localisation,prot$convergence),correct = F)
chisq.test(table(prot$Secondary_Structure,prot$convergence),correct = F)

```

### Structural properties
Here I visualise the structural characteristics of the convergent substitutions to get an sense of where on the protein structure most sites are located.  
Original data found in Datasets/DatasetS0_csv_files/protein_sites/Session_pymol
```{r,fig.dim=c(4,8)}
str_dat<-tibble(surface = c("Surface","Burried"),
                no = c(159,22),
                n.per = round((no/sum(no))*100,digits = 1))
conf_dat<-tibble(conf = c("Helix","Sheet","Loop"),
                 no = c(83,33,65),
                 n.per = round((no/sum(no))*100,digits = 1))


labs1<-paste0(str_dat[[1]], " (", str_dat$n.per, "%)")
labs2<-paste0(conf_dat[[1]], " (", conf_dat$n.per, "%)")
  

g1<-ggpubr::ggbarplot(str_dat,x = "surface",y = "n.per",palette = "npg",color = "black",fill = "surface",label = labs1,xlab = "Localisation of convergnt residues on protein",ylab = "Percentage of convergent sites")
g2<-ggpubr::ggbarplot(conf_dat,x = "conf",y = "n.per",palette = "npg",color = "black",fill = "conf",label = labs2,xlab = "Localisation of convergent residues on secondary structure",ylab = "Percentage of convergent sites")

ggpubr::ggarrange(g1,g2,nrow = 2,legend = "none")

```

### Amino acid property changes
```{r, fig.dim=c(6,5)}
#Load data
Properties_pval_output_table<-readRDS("./RDS/Properties_pval_output_will_test.rds")

#Make plots
Properties_pval_output_table %>% 
  mutate(is_sig=case_when(wil_pval < 0.05 ~ "yes", .default = "no")) %>% 
  filter(is_sig == "yes") %>% group_by(property, state) %>% dplyr::count(is_sig) %>% ungroup() %>% 
  mutate(property=reorder(property,-n)) %>% dplyr::rename("Type of substitution"="state") %>% 
  ggplot(aes(y=property, x=n))+
  geom_col(aes(fill=`Type of substitution`), alpha=0.7, color="black",position = "dodge")+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  xlab("Number of significant changes\nto amino acid property")+
  ylab("Physiochemical properties of amino acids")+
  theme_bw()

Properties_pval_output_table %>% 
  mutate(is_sig=case_when(wil_pval < 0.05 ~ "yes", .default = "no")) %>% 
  filter(is_sig == "yes") %>% group_by(property, state) %>% dplyr::count(is_sig) %>% ungroup() %>% 
  mutate(property=reorder(property,-n)) %>% dplyr::rename("Type of substitution"="state") %>% 
  ggplot(aes(y=property, x=n))+
  geom_col(aes(fill=`Type of substitution`), alpha=0.7, color="black",position = "fill")+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  xlab("Proportion of significant changes\nto amino acid property")+
  ylab("Physiochemical properties of amino acids")+
  theme_bw()
```

## Analysing distribution of tissue specificity {.tabset}
In contrast to the previous analysis where we checked whether convergent genes were more tissue specific than other genes in the genome, this analysis compares the distributions of the values of the tissues-specificity index between convergent genes and all other genes. The distribution of tissue specificity of the convergent genes was significantly different compared to the background (two-sided Kolmogorov-Smirnov test P < 0.05), with the convergent genes having higher distribution at intermediate values of tissue specificity.

### Setting up analysis
```{r, warning=F}
#Load data
skin<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_skin.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
brain<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_brain.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
eye<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_eye.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
heart<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_heart.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
liver<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_liver.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
muscle<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_muscle.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
ovary<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_ovary.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)
testis<-orth_list %>% filter(orths %in% (readRDS("./RDS/Tau_0.8_TMM_cpm_with_log_no_combat_testis.rds") 
                           %>% purrr::list_rbind() %>% dplyr::select(Gene.ID))$Gene.ID)

#Get the orthologs that are exclusively tissue specific for each tissue
#brain
ts_list<-list(brain,eye,heart,liver,muscle,skin,ovary,testis) #change the first entry in the list when computing exclusive orthologs
brain_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Brain_exclusive_orths.csv")
#eye
ts_list<-list(eye,brain,heart,liver,muscle,skin,ovary,testis) #change the first entry in the list when computing exclusive orthologs
eye_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Eye_exclusive_orths.csv")
#heart
ts_list<-list(heart,brain,eye,liver,muscle,skin,ovary,testis) #change the first entry in the list when computing exclusive orthologs
heart_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Heart_exclusive_orths.csv")
#liver
ts_list<-list(liver,eye,brain,heart,muscle,skin,ovary,testis) #change the first entry in the list when computing exclusive orthologs
liver_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Liver_exclusive_orths.csv")
#muscle
ts_list<-list(muscle,eye,brain,heart,liver,skin,ovary,testis) #change the first entry in the list when computing exclusive orthologs
muscle_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Muscle_exclusive_orths.csv")
#skin
ts_list<-list(skin,eye,brain,heart,liver,muscle,ovary,testis) #change the first entry in the list when computing exclusive orthologs
skin_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Skin_exclusive_orths.csv")
#ovary
ts_list<-list(ovary,eye,brain,heart,liver,muscle,skin,testis) #change the first entry in the list when computing exclusive orthologs
ovary_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Overy_exclusive_orths.csv")
#testis
ts_list<-list(testis,eye,brain,heart,liver,muscle,skin,ovary) #change the first entry in the list when computing exclusive orthologs
testis_excl<-purrr::reduce(ts_list,anti_join, by = "Orthogroup")# %>% write_csv("./Datasets/DatasetS0_csv_files/Testis_exclusive_orths.csv")

rbind(brain_excl,eye_excl,heart_excl,liver_excl,muscle_excl,skin_excl,ovary_excl,testis_excl)->excl_tissue_spec

#convergent genes that are also tissue specific
d<-combined_arity %>% filter(Orthogroup %in% non_spuroius_conv$Orthogroup) %>% distinct(Orthogroup)

d<-d %>% mutate(c = "TRUE") %>% 
  mutate(brain_c = if_else(d$Orthogroup %in% brain_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(eye_c = if_else(d$Orthogroup %in% eye_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(heart_c = if_else(d$Orthogroup %in% heart_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(liver_c = if_else(d$Orthogroup %in% liver_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(muscle_c = if_else(d$Orthogroup %in% muscle_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(skin_c = if_else(d$Orthogroup %in% skin_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(ovary_c = if_else(d$Orthogroup %in% ovary_excl$Orthogroup, TRUE, FALSE)) %>% 
  mutate(testis_c = if_else(d$Orthogroup %in% testis_excl$Orthogroup, TRUE, FALSE))


d %>% mutate(spec_in_tissue=
               case_when(brain_c==TRUE ~ "brain",
                         eye_c==TRUE ~ "eye",
                         heart_c==TRUE ~ "heart",
                         liver_c==TRUE ~ "liver",
                         muscle_c==TRUE ~ "muscle",
                         skin_c==TRUE ~ "skin",
                         ovary_c==TRUE ~ "ovary",
                         testis_c==TRUE ~ "testis")) %>% dplyr::select(Orthogroup,spec_in_tissue) %>% drop_na()->excl_conv_tissue

left_join(tau_dat_all$Danio_rerio,
          danio_conv %>% dplyr::select(Orthogroup,orths) %>% dplyr::rename("Gene.ID"="orths") %>% mutate(conv="yes"),
          by = "Gene.ID") %>% distinct(Gene.ID,.keep_all = T) %>% replace_na(list(conv="no")) %>% rowwise() %>%
  mutate(num_tissues = sum(c_across(-c(Gene.ID,tau,conv,,Orthogroup)) > 0.1)) %>%
  ungroup()->Danio_tau_conv_all

left_join(Danio_tau_conv_all, excl_conv_tissue, by = "Orthogroup") %>% mutate(spec_in_tissue=replace_na(spec_in_tissue,"no_tsp")) %>% 
  mutate(is_tsp=case_when(spec_in_tissue=="no_tsp" ~ "no", .default = "yes"))->estimated_ts_conv_Danio_tau_conv_all
```

### Distribution of tissue-specificity
```{r,warning=F,eval=F}
#load all tissue-specificity data
tau_dat_all<-readRDS("./RDS/tau_dat_all.rds")
#Function to plot tissue specificity and compare distributions. 
ts_dis_plots<-function(dat,sp_name){
  inner_join(
    dat,
    orth_list %>% filter(sp==sp_name) %>% dplyr::rename("Gene.ID"=orths) %>% 
      dplyr::select(Gene.ID,Orthogroup), by = "Gene.ID") %>%
    mutate(state=case_when(Orthogroup %in% non_spuroius_conv$Orthogroup ~ "Convergent", .default = "Non-convergent")) %>% mutate(is_tsp=case_when(tau >= 0.8~"yes",.default = "no"))->tau_dist_dat

  #gets p-val for the ks test
  ks_test_pval<-(ks.test(x=tau_dist_dat %>% filter(state=="Convergent") %>% pull(tau),
                         y=tau_dist_dat %>% filter(state=="Non-convergent") %>% pull(tau)))$p.value %>% round(4)
#Plot  
  tau_dist_dat %>% 
    ggplot(aes(x=tau, fill = state, y=..density..))+
    geom_density(alpha=0.6)+
    facet_grid(rows=vars(state))+
    scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
    geom_vline(xintercept = 0.3, linetype="dashed")+#low tissue-specificity
    geom_vline(xintercept = 0.8, linetype="dashed")+#high tissue-specificity
    theme_bw()+
    labs(title=paste0(str_replace(pattern = "_", string = sp_name, replacement = " ")),
         subtitle = paste0("Two-sided KS test p-value: ",ks_test_pval))+
    theme(plot.title = element_text(face = "italic"))->dist_ts
  
  ggpubr::ggarrange(dist_ts,legend = "none") %>% return()
  #tau_dist_dat %>% return() #for dist tests
}

ggpubr::ggarrange(
  ts_dis_plots(tau_dat_all$Astyanax_mexicanus, sp_name = "Astyanax_mexicanus"),
  ts_dis_plots(tau_dat_all$Astatotilapia_calliptera, sp_name = "Astatotilapia_calliptera"),
  ts_dis_plots(tau_dat_all$Danio_rerio, sp_name = "Danio_rerio"),
  ts_dis_plots(tau_dat_all$Esox_lucius, sp_name = "Esox_lucius"),
  ts_dis_plots(tau_dat_all$Gasterosteus_aculeatus, sp_name = "Gasterosteus_aculeatus"),
  ts_dis_plots(tau_dat_all$Gadus_morhua, sp_name = "Gadus_morhua"),
  ts_dis_plots(tau_dat_all$Neolamprologus_brichardi, sp_name = "Neolamprologus_brichardi"),
  ts_dis_plots(tau_dat_all$Nothobranchius_furzeri, sp_name = "Nothobranchius_furzeri"),
  ts_dis_plots(tau_dat_all$Oryzias_latipes, sp_name = "Oryzias_latipes"),
  ts_dis_plots(tau_dat_all$Salmo_salar, sp_name = "Salmo_salar"),
  ts_dis_plots(tau_dat_all$Scophthalmus_maximus, sp_name = "Scophthalmus_maximus"))->ts_dis_dat
#ts_dis_dat %>% saveRDS("./RDS/ts_dis_dat.rds")
```

```{r,fig.dim=c(12,10)}
readRDS("./RDS/ts_dis_dat.rds")
```

### Test of unimodality
Using the Sileverman (Silverman 1981) and Hartigan (Hartigan and Hartigan 1985) tests of unimodality we observed that the tissue-specificity index (tau) of the convergent gene sets resembled a more unimodal distribution as compared to the bimodal distribution of the background set. In this test we set the null hypothesis as the modality we want to test, and check whether the test distribution deviates from it.  For example, we set the null hypothesis as a unimodal distribution, and we check whether the distribution of tau of convergent gene deviates from it. A non-significant p-value indicates we cannot reject the null hypothesis, i.e. our distribution is unimodal. We also compared the distribution of tau for all the other genes and found that they always deviate the null distribution i.e. p-values < 0.05. 

```{r,eval=FALSE}
#function to carry out modality analysis
mode_test<-function(dat,sp){
  library(multimode)
  set.seed(0000) #set seed for the boot strapping
  #Hardigan's dip test
  tibble(sp=sp,
         #this is for the convergent gene sets
         test_sil_pval=(modetest(ts_dis_plots(dat, sp_name = sp) %>% 
                                   filter(Orthogroup %in% non_spuroius_conv$Orthogroup) %>% 
                                   pull(tau), mod0 = 1, method = "HH",B=1000, submethod=NULL))$p.value,
         #this is for the background genes
         bg_sil_pval=(modetest(ts_dis_plots(tau_dat_all$Danio_rerio, sp_name = "Danio_rerio") %>% 
                                 filter(!Orthogroup %in% non_spuroius_conv$Orthogroup) %>% 
                                 pull(tau), mod0 = 1, method = "HH",B=1000, submethod=NULL))$p.value) %>% 
    mutate(is_unimodal=case_when(test_sil_pval > 0.05  & bg_sil_pval < 0.05 ~ "yes", .default = "no"),
           test="Hardigan")->HH
  #Silverman's mode estimator
  tibble(sp=sp,
         test_sil_pval=(modetest(ts_dis_plots(dat, sp_name = sp) %>% 
                                   filter(Orthogroup %in% non_spuroius_conv$Orthogroup) %>% 
                                   pull(tau), mod0 = 1, method = "SI",B=1000, submethod=1))$p.value,
         bg_sil_pval=(modetest(ts_dis_plots(tau_dat_all$Danio_rerio, sp_name = "Danio_rerio") %>% 
                                 filter(!Orthogroup %in% non_spuroius_conv$Orthogroup) %>% 
                                 pull(tau), mod0 = 1, method = "SI",B=1000, submethod=1))$p.value) %>% 
    mutate(is_unimodal=case_when(test_sil_pval > 0.05  & bg_sil_pval < 0.05 ~ "yes", .default = "no"),
           test="Silverman")->SI
  
  rbind(HH,SI) %>% return()
}

rbind(
  mode_test(dat=tau_dat_all$Astatotilapia_calliptera, sp="Astatotilapia_calliptera"),
  mode_test(dat=tau_dat_all$Astyanax_mexicanus, sp="Astyanax_mexicanus"),
  mode_test(dat=tau_dat_all$Danio_rerio, sp="Danio_rerio"),
  mode_test(dat=tau_dat_all$Esox_lucius, sp="Esox_lucius"),
  mode_test(dat=tau_dat_all$Gasterosteus_aculeatus, sp="Gasterosteus_aculeatus"),
  mode_test(dat=tau_dat_all$Gadus_morhua, sp="Gadus_morhua"),
  mode_test(dat=tau_dat_all$Neolamprologus_brichardi, sp="Neolamprologus_brichardi"),
  mode_test(dat=tau_dat_all$Nothobranchius_furzeri, sp="Nothobranchius_furzeri"),
  mode_test(dat=tau_dat_all$Oryzias_latipes, sp="Oryzias_latipes"),
  mode_test(dat=tau_dat_all$Salmo_salar, sp="Salmo_salar"),
  mode_test(dat=tau_dat_all$Scophthalmus_maximus, sp="Scophthalmus_maximus")
)->dist_test_out

```
### Plot the modality test results
```{r,fig.dim=c(3,10)}
dist_test_out<-readRDS("./RDS/dist_test_out.rds")

dist_test_out

dist_test_out %>% 
  ggplot(aes(x=1,y=1, fill=is_unimodal))+
  facet_grid(rows=vars(sp),cols=vars(test))+
  geom_tile(alpha=0.85)+
  scale_fill_manual(values = c("#C78BDD","#47932C"))+
  theme_bw()+
  ylab(NULL)+
  xlab(NULL)+
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x  = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )
```

### Get single cell distribution
In this analysis I look at the distribution of convergent genes and the different cells types they are expressed in. The rationale is to check whether on an average, the convergent genes are expressed in different number of cell types as compared to a background set of genes. Here, the background is all other genes, resulting in a null distribution. Using the Kolmogorov-Smirnov test, I test for divergence between the two distributions. Using the two-sided test I see that the distributions differ from one another, and using the one-sided test I check whether the values in the convergent group and higher than the test group at corresponding positions along the distributions where they diverge. The KS test shows that the values are greater at the divergent points, suggesting that convergent group of genes on an average are expressed in more cell types.
```{r,eval=F}
library(SingleCellExperiment)
library(scater)
library(scuttle)

danio_conv<-read_csv("./Datasets/DatasetS0_csv_files/Danio_rerio_convergent_genes.csv")
Danio_biomart<-read_csv("./Datasets/DatasetS0_csv_files/Danio_biomart.csv")
zhub_dat<-readRDS("./RDS/zhub_full_log.rds")

left_join(estimated_ts_conv_Danio_tau_conv_all,
          Danio_biomart %>% dplyr::select(`Gene stable ID`, `Gene name`) %>% 
            dplyr::rename("Gene.ID"=`Gene stable ID`),
          by = "Gene.ID") %>% distinct(Gene.ID, .keep_all = T)->estimated_ts_conv_Danio_tau_conv_all

#Function to load single cell data and process it
single_cell_df<-function(stage,gene_in){
  
  zhub_dat<-readRDS(paste0("./RDS/zhub_",stage,"_log.rds"))
  scuttle::makePerCellDF(zhub_dat, features = gene_in, assay.type = "logcounts", swap.rownames="gene_ids")  %>% 
  dplyr::rename("logCounts"=gene_in) %>% filter(logCounts >= 0.5) %>% mutate(gene=gene_in)->cell_df
  
  num_of_samples<-(length(cell_df %>% distinct(fish) %>% pull(fish))-1)
  
  #Keep the cells with expression in at least max-1 samples
    cell_df %>% group_by(zebrafish_anatomy_ontology_class) %>% summarise(count_fish=n_distinct(fish), .groups = "keep") %>% 
      ungroup() %>% filter(count_fish >= num_of_samples)->filter_in_cell_type
  
  cell_df %>% filter(zebrafish_anatomy_ontology_class %in% filter_in_cell_type$zebrafish_anatomy_ontology_class) %>% 
    dplyr::select(gene,logCounts,timepoint,zebrafish_anatomy_ontology_class,
                  zebrafish_anatomy_ontology_id,fish,
                  X_umap.1,X_umap.2)->cell_df
  
    return(cell_df)
}


single_cell_df<-possibly(.f = single_cell_df, otherwise = NULL)


purrr::map(.x=as.list(head(estimated_ts_conv_Danio_tau_conv_all %>% 
    filter(conv=="yes") %>% pull(Gene.ID))),
           ~single_cell_df(stage = "10hpf",gene_in = .x)) %>% list_rbind()->conv_cells_10hpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "12hpf",gene_in = .x)) %>% list_rbind()->conv_cells_12hpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "14hpf",gene_in = .x)) %>% list_rbind()->conv_cells_14hpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "16hpf",gene_in = .x)) %>% list_rbind()->conv_cells_16hpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "19hpf",gene_in = .x)) %>% list_rbind()->conv_cells_19hpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "24hpf",gene_in = .x)) %>% list_rbind()->conv_cells_24hpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "3dpf",gene_in = .x)) %>% list_rbind()->conv_cells_3dpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "5dpf",gene_in = .x)) %>% list_rbind()->conv_cells_5dpf

purrr::map(.x=as.list(estimated_ts_conv_Danio_tau_conv_all %>% 
                        filter(conv=="yes") %>% pull(Gene.ID)),
           ~single_cell_df(stage = "10dpf",gene_in = .x)) %>% list_rbind()->conv_cells_10dpf


#All cells
system.time(
purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids[1:20]),
           ~single_cell_df(stage = "10hpf",gene_in = .x)) %>% list_rbind()->all_gene_10hpf
)


purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "12hpf",gene_in = .x)) %>% list_rbind()->all_gene_12hpf
all_gene_12hpf %>% saveRDS("./RDS/all_gene_12hpf.rds")

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "14hpf",gene_in = .x)) %>% list_rbind()->all_gene_14hpf
all_gene_14hpf %>% saveRDS("./RDS/all_gene_14hpf.rds")

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "16hpf",gene_in = .x)) %>% list_rbind()->all_gene_16hpf

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "19hpf",gene_in = .x)) %>% list_rbind()->all_gene_19hpf

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "24hpf",gene_in = .x)) %>% list_rbind()->all_gene_24hpf

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "3dpf",gene_in = .x)) %>% list_rbind()->conv_cells_3dpf

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "5dpf",gene_in = .x)) %>% list_rbind()->all_gene_5dpf

purrr::map(.x=as.list(rowData(zhub_dat)$gene_ids),
           ~single_cell_df(stage = "10dpf",gene_in = .x)) %>% list_rbind()->all_gene_10dpf

sc_dist_analysis<-function(sc_dat, plot_title){

#count the number of unique cell types per gene and get their ortholog annotations
inner_join(
  sc_dat %>% summarise(no_uniq_cells_types=n_distinct(zebrafish_anatomy_ontology_class), .by = "gene"),
  orth_list %>% dplyr::select(Orthogroup,orths,sp) %>% filter(sp=="Danio_rerio") %>% dplyr::rename("gene"=orths),by = "gene") %>% dplyr::rename("Gene.ID"=gene)->cell_types_orth
  
#Get information on convergence
rbind(
cell_types_orth %>% filter(Orthogroup %in% non_spuroius_conv$Orthogroup) %>% mutate(state="conv"),
cell_types_orth %>% filter(!Orthogroup %in% non_spuroius_conv$Orthogroup) %>% mutate(state="non_conv"))->test_dat
#Get information of tissue specificity
test_dat %>% mutate(is_tsp=case_when(Orthogroup %in% (estimated_ts_conv_Danio_tau_conv_all %>% filter(is_tsp=="yes"))$Orthogroup ~ "yes",.default = "no"))->test_dat

(ks.test(x=test_dat  %>% filter(state=="conv") %>% pull(no_uniq_cells_types),
        y=test_dat  %>% filter(state=="non_conv") %>% pull(no_uniq_cells_types),
            alternative = "two.sided"))$p.value->ks_pval_two_sided
(ks.test(x=test_dat  %>% filter(state=="conv") %>% pull(no_uniq_cells_types),
        y=test_dat  %>% filter(state=="non_conv") %>% pull(no_uniq_cells_types),
            alternative = "greater"))$p.value->ks_pval_greater
(ks.test(x=test_dat  %>% filter(state=="conv") %>% pull(no_uniq_cells_types),
        y=test_dat  %>% filter(state=="non_conv") %>% pull(no_uniq_cells_types),
            alternative = "less"))$p.value->ks_pval_less

test_dat %>% 
  ggplot(aes(x=no_uniq_cells_types, fill = state))+
  geom_density(alpha=0.6)+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  xlab("Number of unique cell types")+
  labs(title = str_remove(string = paste0(plot_title), 
                          pattern ="all_gene_"),
       subtitle = paste0("One-sided KS-test p-value: ", signif(ks_pval_greater, digits = 3)))+
  theme_bw()+
  theme(text = element_text(size=8))->dist_plot

test_dat  %>%
  ggplot(aes(x=no_uniq_cells_types, color= state))+
  stat_ecdf()+
  xlab("Number of unique cell types")+
  scale_color_manual(values = c("#EA5B15","#1AAEC0"))+
  theme_bw()+
  labs(title = "Cumulative probability distrubutions")+
  theme(text = element_text(size=8))->ecdf_plot

ggarrange(dist_plot, ecdf_plot, ncol = 1, common.legend = T)->c_plot


return(list(dist_plot_dat=dist_plot$data,c_plot=c_plot, ks_pval_two_sided=ks_pval_two_sided, ks_pval_greater=ks_pval_greater, ks_pval_less=ks_pval_less))
}

#Load processed sc data frame and run above function
all_gene_10hpf<-readRDS("./RDS/all_gene_10hpf.rds")
sc_dist_analysis(sc_dat = all_gene_10hpf, plot_title= "Distribution of gene expression\nin unique cell types at 10hpf")->dist_res_10hpf
dist_res_10hpf %>% saveRDS("./RDS/dist_res_10hpf.rds")
rm(all_gene_10hpf)

all_gene_12hpf<-readRDS("./RDS/all_gene_12hpf.rds")
sc_dist_analysis(sc_dat = all_gene_12hpf, plot_title= "Distribution of gene expression\nin unique cell types at 12hpf")->dist_res_12hpf
dist_res_12hpf %>% saveRDS("./RDS/dist_res_12hpf.rds")
rm(all_gene_12hpf)

all_gene_14hpf<-readRDS("./RDS/all_gene_14hpf.rds")
sc_dist_analysis(sc_dat = all_gene_14hpf, plot_title= "Distribution of gene expression\nin unique cell types at 14hpf")->dist_res_14hpf
dist_res_14hpf %>% saveRDS("./RDS/dist_res_14hpf.rds")
rm(all_gene_14hpf)

all_gene_16hpf<-readRDS("./RDS/all_gene_16hpf.rds")
sc_dist_analysis(sc_dat = all_gene_16hpf, plot_title= "Distribution of gene expression\nin unique cell types at 16hpf")->dist_res_16hpf
dist_res_16hpf %>% saveRDS("./RDS/dist_res_16hpf.rds")
rm(all_gene_16hpf)

all_gene_19hpf<-readRDS("./RDS/all_gene_19hpf.rds")
sc_dist_analysis(sc_dat = all_gene_19hpf, plot_title= "Distribution of gene expression\nin unique cell types at 19hpf")->dist_res_19hpf
dist_res_19hpf %>% saveRDS("./RDS/dist_res_19hpf.rds")
rm(all_gene_19hpf)

all_gene_24hpf<-readRDS("./RDS/all_gene_24hpf.rds")
sc_dist_analysis(sc_dat = all_gene_24hpf, plot_title= "Distribution of gene expression\nin unique cell types at 24hpf")->dist_res_24hpf
dist_res_24hpf %>% saveRDS("./RDS/dist_res_24hpf.rds")
rm(all_gene_24hpf)

all_gene_3dpf<-readRDS("./RDS/all_gene_3dpf.rds")
sc_dist_analysis(sc_dat = all_gene_3dpf, plot_title= "Distribution of gene expression\nin unique cell types at 3dpf")->dist_res_3dpf
dist_res_3dpf %>% saveRDS("./RDS/dist_res_3dpf.rds")
rm(all_gene_3dpf)

all_gene_5dpf<-readRDS("./RDS/all_gene_5dpf.rds")
sc_dist_analysis(sc_dat = all_gene_5dpf, plot_title= "Distribution of gene expression\nin unique cell types at 5dpf")->dist_res_5dpf
dist_res_5dpf %>% saveRDS("./RDS/dist_res_5dpf.rds")
rm(all_gene_5dpf)

all_gene_10dpf<-readRDS("./RDS/all_gene_10dpf.rds")
sc_dist_analysis(sc_dat = all_gene_10dpf, plot_title= "Distribution of gene expression\nin unique cell types at 10dpf")->dist_res_10dpf
dist_res_10dpf %>% saveRDS("./RDS/dist_res_10dpf.rds")#This takes very long; couple of hours and is a very large rds (~1.5GB). Therefore, have not included
rm(all_gene_10dpf)

dist_res_10hpf<-readRDS("./RDS/dist_res_10hpf.rds")
dist_res_12hpf<-readRDS("./RDS/dist_res_12hpf.rds")
dist_res_14hpf<-readRDS("./RDS/dist_res_14hpf.rds")
dist_res_16hpf<-readRDS("./RDS/dist_res_16hpf.rds")
dist_res_19hpf<-readRDS("./RDS/dist_res_19hpf.rds")
dist_res_24hpf<-readRDS("./RDS/dist_res_24hpf.rds")
dist_res_3dpf<-readRDS("./RDS/dist_res_3dpf.rds")
dist_res_5dpf<-readRDS("./RDS/dist_res_5dpf.rds")
dist_res_10dpf<-readRDS("./RDS/dist_res_10dpf.rds")#This takes very long; couple of hours and is a very large rds (~1.5GB). Therefore, have not included
```


```{r,eval=FALSE}
dist_res_10hpf
```

### Running cell distribution analysis for random set of genes
Here I run the analysis using a random set of genes. While visually the distributions may look similar at times, the KS test shows that their distributions are not different from the background. This suggests that the convergent genes have a unique distribution of gene expression in different cell types as opposed to any random set of genes. 
I am showing the results here for 10hpf and 10dps, Users can run the code for any other time point. 
```{r,eval=FALSE}
sc_dist_random<-function(sc_dat, plot_title){

#count the number of unique cell types per gene and get their ortholog annotations
inner_join(
  sc_dat %>% summarise(no_uniq_cells_types=n_distinct(zebrafish_anatomy_ontology_class), .by = "gene"),
  orth_list %>% dplyr::select(Orthogroup,orths,sp) %>% filter(sp=="Danio_rerio") %>% dplyr::rename("gene"=orths),by = "gene") %>% dplyr::rename("Gene.ID"=gene)->cell_types_orth
  
#Takes a random set of 89 orthogroups 
rbind(
  cell_types_orth %>% filter(Orthogroup %in% (orth_list %>% distinct(Orthogroup) %>% pull(Orthogroup))[runif(89, min = 1, max = 8308)]) %>% 
    mutate(state="conv"), cell_types_orth %>% filter(!Orthogroup %in% non_spuroius_conv$Orthogroup) %>% mutate(state="non_conv"))->test_dat

#Get information of tissue specificity
test_dat %>% mutate(is_tsp=case_when(Orthogroup %in% (estimated_ts_conv_Danio_tau_conv_all %>% filter(is_tsp=="yes"))$Orthogroup ~ "yes",.default = "no"))->test_dat


(ks.test(x=test_dat  %>% filter(state=="conv") %>% pull(no_uniq_cells_types),
        y=test_dat  %>% filter(state=="non_conv") %>% pull(no_uniq_cells_types),
            alternative = "two.sided"))$p.value->ks_pval_two_sided
(ks.test(x=test_dat  %>% filter(state=="conv") %>% pull(no_uniq_cells_types),
        y=test_dat  %>% filter(state=="non_conv") %>% pull(no_uniq_cells_types),
            alternative = "greater"))$p.value->ks_pval_greater
(ks.test(x=test_dat  %>% filter(state=="conv") %>% pull(no_uniq_cells_types),
        y=test_dat  %>% filter(state=="non_conv") %>% pull(no_uniq_cells_types),
            alternative = "less"))$p.value->ks_pval_less

test_dat %>% 
  ggplot(aes(x=no_uniq_cells_types, fill = state))+
  geom_density(alpha=0.6)+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  xlab("Number of unique cell types")+
  labs(title = str_remove(string = paste0(plot_title), 
                          pattern ="all_gene_"),
       subtitle = paste0("One-sided KS-test p-value: ", signif(ks_pval_greater, digits = 3)))+
  theme_bw()+
  theme(text = element_text(size=8))->dist_plot

test_dat  %>%
  ggplot(aes(x=no_uniq_cells_types, color= state))+
  stat_ecdf()+
  xlab("Number of unique cell types")+
  scale_color_manual(values = c("#EA5B15","#1AAEC0"))+
  theme_bw()+
  labs(title = "Cumulative probability distrubutions")+
  theme(text = element_text(size=8))->ecdf_plot


ggarrange(dist_plot, ecdf_plot, ncol = 1, common.legend = T)->c_plot

return(list(c_plot=c_plot, ks_pval_two_sided=ks_pval_two_sided, ks_pval_greater=ks_pval_greater, ks_pval_less=ks_pval_less))
}

sc_dist_random(sc_dat = all_gene_10hpf, plot_title= "Random genes at 10hpf")->dist_random_10hpf

sc_dist_random(sc_dat = all_gene_10dpf, plot_title= "Random genes at 10dpf")->dist_random_10dpf

```

## Sementic Similaritry of gene sets {.tabset}
In this section I check whether the convergent genes and sets of non-convergent (background) genes differ in terms of the diversity of biological features they are associated with. To do this I estimate Sementic Similaritry (SS) of the GO terms. SS uses the information content of GO terms as well as their ontology structure to allow comparison of genes annotated with GO terms. Higher the SS the more closely related two gene products (or gene sets are). I estimate the pairwise SS of two gene sets, the convergent genes and a background gene set. I found that the convergent genes on an average, have lower SS for Biological Process terms and Cellular Component terms than a background set (one-sided KS-test). This implies that the convergent gene sets are likely associate with more diverse set of processes and are expressed in a more diverse cellular structures than the background. 

### Loading the GO annotations
```{r, eval=F}
library(GOSemSim)
library(org.Dr.eg.db)
library(AnnotationDbi)

#Load Danio data
read_csv("./Datasets/DatasetS0_csv_files/Danio_biomart.csv")->Danio_biomart
danio_conv<-read_csv("./Datasets/DatasetS0_csv_files/Danio_rerio_convergent_genes.csv")
#Get GO data
go_annotations <- select(org.Dr.eg.db,
                         keys = keys(org.Dr.eg.db, keytype = "ENSEMBL"), 
                         columns = c("GO", "ONTOLOGY"), 
                         keytype = "ENSEMBL")

#Prepare the semantic data
drGO_BP<-godata('org.Dr.eg.db', keytype="ENSEMBL", ont="BP")
drGO_MF<-godata('org.Dr.eg.db', keytype="ENSEMBL", ont="MF")
drGO_CC<-godata('org.Dr.eg.db', keytype="ENSEMBL", ont="CC")

#Specify gene sets
geneset_conv<-orth_list %>% filter(sp=="Danio_rerio") %>% filter(Orthogroup %in% non_spuroius_conv$Orthogroup)
geneset_nc<-orth_list %>% filter(sp=="Danio_rerio") %>% filter(!Orthogroup %in% non_spuroius_conv$Orthogroup) 

#Retrieve annotations
go_annotations_BP <- go_annotations %>% filter(ONTOLOGY=="BP")
go_annotations_MF <- go_annotations %>% filter(ONTOLOGY=="MF")
go_annotations_CC <- go_annotations %>% filter(ONTOLOGY=="CC")
```

### Calculating SS
```{r,eval=FALSE}
#Select a random geneset a background
#I've repeated the analysis using multiple random gene sets are the result is always the same. Try!
#Selecting 120 Orthogroups gives around the same number of genes
#We also ran this with 2000 random orthogroups as the background, and we sill get the same results. Note, this analysis can take some time.  
#set.seed(1234)
random_set<-(geneset_nc %>% distinct(Orthogroup))$Orthogroup[runif(120,min = 1,9061) %>% round()] 

#Compute SS between gene sets for each GO term category
mgeneSim(genes=geneset_conv$orths,
         semData=drGO_BP, measure="Wang",verbose=FALSE, combine = "BMA")->gene_sem_conv_BP
mgeneSim(genes=geneset_nc %>% 
           filter(Orthogroup %in% random_set) %>% pull(orths),
         semData=drGO_BP, measure="Wang",verbose=FALSE, combine = "BMA")->gene_sem_nc_BP
mgeneSim(genes=geneset_conv$orths,
         semData=drGO_MF, measure="Wang",verbose=FALSE, combine = "BMA")->gene_sem_conv_MF
mgeneSim(genes=geneset_nc %>% 
           filter(Orthogroup %in% random_set) %>% pull(orths),
         semData=drGO_MF, measure="Wang",verbose=FALSE, combine = "BMA")->gene_sem_nc_MF
mgeneSim(genes=geneset_conv$orths,
         semData=drGO_CC, measure="Wang",verbose=FALSE, combine = "BMA")->gene_sem_conv_CC
mgeneSim(genes=geneset_nc %>% 
           filter(Orthogroup %in% random_set) %>% pull(orths),
         semData=drGO_CC, measure="Wang",verbose=FALSE, combine = "BMA")->gene_sem_nc_CC

#remove self comparisons and symetric comparisons and combine into one df
rbind(
  reshape2::melt(gene_sem_nc_BP) %>% filter(Var1!=Var2) %>%
    distinct(Var1, Var2, .keep_all = TRUE)%>% mutate(type="Non-convergent",GO_group="BP"),
  reshape2::melt(gene_sem_nc_MF) %>% filter(Var1!=Var2) %>%
    distinct(Var1, Var2, .keep_all = TRUE)%>% mutate(type="Non-convergent",GO_group="MF"),
  reshape2::melt(gene_sem_nc_CC) %>% filter(Var1!=Var2) %>%
    distinct(Var1, Var2, .keep_all = TRUE)%>% mutate(type="Non-convergent",GO_group="CC"),
  reshape2::melt(gene_sem_conv_BP) %>% filter(Var1!=Var2) %>%
    distinct(Var1, Var2, .keep_all = TRUE)%>% mutate(type="Convergent",GO_group="BP"),
  reshape2::melt(gene_sem_conv_MF) %>% filter(Var1!=Var2) %>%
    distinct(Var1, Var2, .keep_all = TRUE)%>% mutate(type="Convergent",GO_group="MF"),
  reshape2::melt(gene_sem_conv_CC) %>% filter(Var1!=Var2) %>%
    distinct(Var1, Var2, .keep_all = TRUE)%>% mutate(type="Convergent",GO_group="CC"))->ss_test


ss_test %>% saveRDS("./RDS/ss_test_Wang.rds")
#ss_test %>% saveRDS("./RDS/ss_test_Wang_2k.rds") #Another data set with 2 orths. Gives same result. 
```

```{r, fig.dim=c(3,8)}
ss_test<-readRDS("./RDS/ss_test_Wang.rds")
#Plot the difference in SS between groups with the test p-value; 
#the one-sided test checks whether the divergence of the convergent distribution is significantly higher at the corresponding values. 
ss_test %>% filter(GO_group=="BP") %>% 
  ggplot(aes(x=log2(value), fill = type))+
  geom_density(color="black", alpha=0.6)+
  theme_bw()+  
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  xlab("log2(Wang's sementic similarity)")+
  labs(title = "Biological Process Terms",
       subtitle = paste0("One-sided KS p-value = ",
                         signif(ks.test(ss_test %>% filter(GO_group=="BP") %>% filter(type=="Convergent") %>% pull(value),
                                            ss_test %>% filter(GO_group=="BP") %>% filter(type=="Non-convergent") %>% pull(value),
                                            alternative = "greater", exact = T)$p.value,digits = 3)))->g_bp_ks

ss_test %>% filter(GO_group=="MF") %>% 
  ggplot(aes(x=log2(value), fill = type))+
  geom_density(color="black", alpha=0.6)+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  theme_bw()+  
  xlab("log2(Wang's sementic similarity)")+
  labs(title = "Molecular Function Terms",
       subtitle = paste0("One-sided KS p-value = ",
                         signif(ks.test(ss_test %>% filter(GO_group=="MF") %>% filter(type=="Convergent") %>% pull(value),
                                            ss_test %>% filter(GO_group=="MF") %>% filter(type=="Non-convergent") %>% pull(value),
                                            alternative = "greater",exact = T)$p.value, digits = 3)))->g_mf_ks


ss_test %>% filter(GO_group=="CC") %>% 
  ggplot(aes(x=log2(value), fill = type))+
  geom_density(color="black", alpha=0.6)+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  theme_bw()+  
  xlab("log2(Wang's sementic similarity)")+
  labs(title = "Cellular Component Terms",
       subtitle = paste0("One-sided KS test p-value = ",
                         signif(ks.test(ss_test %>% filter(GO_group=="CC") %>% filter(type=="Convergent") %>% pull(value),
                                            ss_test %>% filter(GO_group=="CC") %>% filter(type=="Non-convergent") %>% pull(value),
                                            alternative = "greater",exact = T)$p.value, digits = 3)))->g_cc_ks

ggpubr::ggarrange(g_bp_ks,g_mf_ks,g_cc_ks,ncol = 1, common.legend = T,legend = "none")
```

### SS Dendrogram
In this section I do a semantic similarity analysis between GO terms and then make a dendrogram to visualise their diversity and similarity. 
```{r,eval=F}
##Compute SS between terms
SemSim_compute<-function(ensembl,annot,measure, semData){
expand_grid(c1=annot %>% filter(ENSEMBL==paste0(ensembl)) %>% pull(GO),
            c2=annot %>% filter(ENSEMBL==paste0(ensembl)) %>% pull(GO)) %>% 
    filter(c1 != c2) %>% 
  # Ensure symmetric rows are removed
  rowwise() %>%
  mutate(pair = list(sort(c(c1, c2)))) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  dplyr::select(-pair)->dat

#Make the function to calculate semantic similarity
avg_semsim<-function(go1,go2,measure,combine,ensembl){
  goSim(go1, go2, semData=semData, measure=measure)->res
  
  data.frame(GO1=go1,
             GO2=go2,
             ss=res,
             ensembl_id=ensembl)->res_dat
  
  return(res_dat)
}

purrr::pmap(list(.x = as.list(dat$c1),
                 .y = as.list(dat$c2)), 
            ~avg_semsim(go1=.x, go2=.y, 
                        ensembl=ensembl, 
                        measure = measure)) %>% list_rbind()->ss_res
return(ss_res)

}
#To get the graphs for the other categories, replace go_annotations_BP, and drGO_BP
purrr::map(.x=as.list(geneset_conv$orths),
           ~SemSim_compute(ensembl = .x, 
                           annot=go_annotations_BP, 
                           semData=drGO_BP, measure="Wang")) %>% list_rbind()->ss_out_conv

# Function to get GO term descriptions
get_go_description <- function(go_terms) {
  library(GO.db)
  map_chr(go_terms, ~ AnnotationDbi::select(GO.db, keys = .x, columns = "TERM", keytype = "GOID")$TERM[1] %||% NA)
}
# Add descriptions to dataframe
df <- ss_out_conv %>% na.omit() %>% 
  mutate(
    GO1_Description = get_go_description(GO1),
    GO2_Description = get_go_description(GO2)
  )
#df %>% write_csv("./Datasets/DatasetS0_csv_files/SS_graph_data_CC.csv")

```

These plots can be found in Datasets S12 in the online repository.
```{r,eval=F}
library(enrichplot)
library(clusterProfiler)
df<-read_csv("./Datasets/DatasetS0_csv_files/SS_graph_data_BP.csv")
#df <- df %>% na.omit() %>% filter(ss <= 0.5) #for the figure in the main text I use this filter to make visualisation clearer. 

# Make custom enrichResult object
data.frame(ID=c(df$GO1,df$GO2)) %>% mutate(Description=get_go_description(c(df$GO1,df$GO2)))->t
inner_join(t,df %>% dplyr::select(GO1,ensembl_id) %>% 
             dplyr::rename("ID"="GO1"), by = "ID",relationship = "many-to-many") %>% 
  distinct(ID,ensembl_id, .keep_all = T)->t
t %>% group_by(ID, Description) %>% mutate(Count=n_distinct(ensembl_id)) %>% ungroup()->t
t %>% mutate(pvalue=0.05, p.adjust = 0.05) %>% dplyr::rename("geneID"="ensembl_id")->t2
t2 %>% distinct(ID, .keep_all = T) %>% column_to_rownames("ID") %>% 
  mutate(ID=t2 %>% distinct(ID) %>% pull(ID)) %>% 
  dplyr::select(ID, Description, pvalue,p.adjust,geneID,Count)->t2

custom_enrich <- new(
  "enrichResult",
  result = t2 %>% distinct(ID, .keep_all = T),
  organism = "custom",
  keytype = "ENSEMBL",
  ontology = "BP",#Can be kept the same
  gene = unlist(strsplit(t2$geneID, "/")),
  geneSets = list(),
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH"
)

go_db <- godata('org.Dr.eg.db', ont="BP", computeIC=FALSE)#Change ont for other categories
sim_matrix <- setReadable(custom_enrich, 'org.Dr.eg.db', 'ENSEMBL')
sim_matrix <- enrichplot::pairwise_termsim(custom_enrich, semData=go_db, method = "Wang")
p_sim_matrix<- enrichplot::treeplot(sim_matrix)


dist_matrix <- as.dist(1 - sim_matrix@termsim) 
hc <- hclust(dist_matrix, method = "ward.D2")

hc_dendro<-as.dendrogram(hc)
hc_dendro_data <- ggdendro::dendro_data(hc_dendro,type="rectangle")

labels_df <- hc_dendro_data$labels
left_join(
  left_join(labels_df,t2 %>% dplyr::select(Description,geneID),by=c("label"="Description")),
  danio_conv %>% dplyr::select(orths,`Gene name`), by=c("geneID"="orths")) %>% ungroup() %>% distinct(label,`Gene name`) %>% mutate(label=paste0(`Gene name`))->labels_gene_name
gene_colors <- colorspace::qualitative_hcl(n = 68, palette = "set3")
tibble(labels_gene_name %>% distinct(`Gene name`),
       gene_colors=gene_colors)
left_join(labels_gene_name,
          tibble(labels_gene_name %>% distinct(`Gene name`),
       gene_colors=gene_colors),
       by=c("Gene name"="Gene name"))->colors_gene

labels_df2<-tibble(x=labels_df$x, y=labels_df$y, label=labels_df$label, label2=labels_gene_name$label, col=colors_gene$gene_colors)

#These labels are for BP; users have to annotate the other categories themselves. 
#labels_df$type<-c(rep("Development",36),
#                  rep("Translation and protein processing",32),
#                  rep("Metabolic process",11),
#                  rep("Transport",20),
#                  rep("Immune response",21),
#                  rep("Pigmentation and behaviour",4),
#                  rep("Mitotic process",15),
#                  rep("Signalling pathways",17),
#                  rep("Reproduction related",3),
#                  rep("Cell cycle and homeostasis",12))


library(ggdendro)
ggplot(segment(hc_dendro_data)) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend))+
  geom_text(data = labels_df2, 
            aes(x=x, y=y-0.05, label=label2, colour = col),
            hjust = 0,vjust = 0.4, size=3,fontface="italic")+
  geom_text(data = labels_df2, 
            aes(x=x, y=y-0.5, label=label),
            hjust = 0,vjust = 0.4, size=3)+
  coord_flip()+
  scale_y_reverse(expand = c(0.9, 0))+
  scale_color_identity()+
  theme_void()

```


## Phenotypic effect distribution {.tabset}
We collected data from the Zebrafish International Network (ZFIN) database to examine the distribution of unique phenotypic terms associated with genetic perturbation (knockdowns, mutation screens, and morpholino experiments) in Danio rerio. The ZFIN database provides a wide array of expertly curated cross referenced genetic and genomics data that is routinely used to study gene functions.  For the convergent genes with data in ZFIN, we observed that genetic perturbation can affect several structures across multiple tissue systems. Compared to the background of all genes with available phenotypic data, the convergent genes had a higher distribution at intermediate levels of affected phenotypic terms, significantly different from the background (one-sided Kolmogorov-Smirnov test P < 0.05)

### Load the data
```{r,eval=FALSE}
read_csv("./Datasets/DatasetS0_csv_files/Danio_biomart.csv")->Danio_biomart #Obtained from bioMart package
read_tsv("./Datasets/DatasetS0_csv_files/phenoGeneCleanData_fish_2023.05.07.tsv")->pheno #Download from ZFIN



get_danio<-function(data,name){
orth_list %>% filter(Species=="Danio_rerio") %>% filter(Orthogroup %in% data)->d

Danio_biomart %>% filter(`Gene stable ID` %in% d$orths) %>% distinct(`Gene name`) %>%  write_csv(paste0("./Datasets/DatasetS0_csv_files/",name,"_convergent_Danio_genes.csv"))
}

get_danio(non_spuroius_conv$Orthogroup,"Non_spurious_conv")

orth_list %>% filter(Orthogroup %in% non_spuroius_conv$Orthogroup) %>% filter(sp == "Danio_rerio") %>% dplyr::select(Orthogroup,orths)->dan


left_join(dan,pheno %>% filter(`Gene Symbol` %in% dan$`Gene name`) %>% dplyr::rename(`Gene name`=`Gene Symbol`), by = "Gene name") %>% write_csv("./Datasets/DatasetS0_csv_files/Danio_rerio_convergent_genes_pheno.csv")
```

### ZFIN table
```{r,echo=FALSE,include=TRUE,message=FALSE}
pheno_conv<-read_csv("./Datasets/DatasetS0_csv_files/Danio_rerio_convergent_genes_pheno.csv")

kbl(head(pheno_conv,1000)) %>% kable_styling() %>%
  scroll_box(width = "1000px", height = "400px")
```

### Phenotype plots
The plot shows the distribution of phenotype keyword terms and the frequency of affected phenotype terms from ZFIN. Panel on the left shows the effect of genetic perturbation on the convergent genes. Panel on the right shows the number of unique affected structures that show a phenotype after genetic perturbation. The data shows that mutations in the convergent genes can affect several phenotypes and structures, with a majority of the effects showing a decrease or reduction in affected structures or processes, while a few shows an increase.
```{r,fig.dim=c(16,12)}
ph_dat<-pheno_conv %>% summarise(n = n(),.by = c(`Phenotype Keyword Name`)) %>% mutate(`Phenotype Keyword Name`=reorder(`Phenotype Keyword Name`,n))
g1<-ggpubr::ggbarplot(ph_dat, y = "Phenotype Keyword Name", x = "n", 
                  x.text.angle = 90,
                  fill = "#708090",
                  color = "white",
                  xlab = "Frequency of keyword")+
    geom_text(aes(label = n), hjust = -0.2)


read_tsv("./Datasets/DatasetS0_csv_files/phenoGeneCleanData_fish_2023.05.07.tsv")->pheno #Download from ZFIN
#ph_c<-pheno %>% filter(`Gene Symbol` %in% pheno_conv$`Gene name`)
ph_c<-pheno %>% filter(`Gene Symbol` %in% pheno_conv$`Gene name`) %>% dplyr::select(`Gene Symbol`,`Affected Structure or Process 1 superterm ID`) %>% 
  na.omit() %>% distinct(`Affected Structure or Process 1 superterm ID`,.keep_all = T) %>% 
  summarise(n = n(),.by = c(`Gene Symbol`)) %>% 
  mutate(type = "convergent") %>% arrange(desc(n))


make_pheno_terms_dist<-function(d){
  pheno %>% filter(`Gene Symbol` == d) %>% distinct(`Affected Structure or Process 1 superterm ID`,.keep_all = T) %>% 
  summarise(n = n(),.by = c(`Affected Structure or Process 1 superterm Name`,`Publication ID`)) %>% mutate(Gene_symbol = d)
}

purrr::map(ph_c$`Gene Symbol`,make_pheno_terms_dist) %>% bind_rows() %>% 
  dplyr::select(Gene_symbol,`Affected Structure or Process 1 superterm Name`) %>% 
  summarise(n = n(),.by = c(Gene_symbol))->ph_c

ph_c$Gene_symbol<-factor(ph_c$Gene_symbol,levels = c(ph_c %>% arrange((n)) %>% pull(Gene_symbol)))
g2<-ggpubr::ggbarplot(ph_c,y = "Gene_symbol", 
                      x = "n",
                      fill = "#708090",
                      color = "white",
                      x.text.angle = 0,
                      xlab = "Frequency of Affected structure superterm",
                      ylab = "Genes")+
  geom_text(aes(label = n), hjust = -0.1)

ggpubr::ggarrange(g1,g2)

```

The chord diagram shows that for some genes the perturbation phenotypes occur across multiple tissue systems (e.g. alcama, pax3a). The length of the grey bars correspond to the frequency values in the bar plot.
```{r}
library(circlize)
pheno_terms<-read_csv("./Datasets/DatasetS0_csv_files/Pheno_terms_table_revised.csv")
pheno_terms %>% summarise(n = n(),.by = c(`Gene Symbol`,tissue_system))

pheno_terms_wide<-pheno_terms %>% summarise(n = n(),.by = c(`Gene Symbol`,tissue_system)) %>% 
  pivot_wider(names_from = `Gene Symbol`, values_from = n) 
pheno_terms_wide[is.na(pheno_terms_wide)]=0

df_matrix <- pheno_terms_wide %>%
  select(-tissue_system) %>%  # Remove the Gene_symbol column
  as.matrix()

rownames(df_matrix)<-pheno_terms_wide$tissue_system

tissue_colors_vibrant <- c(
  "Nervous" = "#FF4500",      # Orange Red
  "Skeletal" = "#1E90FF",     # Dodger Blue
  "Circulatory" = "#9400D3",  # Dark Violet
  "Whole organism" = "#32CD32", # Lime Green
  "Digestive" = "#FF7F50",    # Coral
  "Embryo"= "#2611EE",        # Dark Navy
  "Skin" = "#FFD700",         # Gold
  "Excretory" = "#8B0000",    # Dark Red
  "Reproductive" = "#FF1493", # Deep Pink
  "Cell process" = "#00FFFF", # Cyan
  "Immune" = "#EE82EE"     # Violet
)


chordDiagram(df_matrix, 
             annotationTrack = "grid", 
             preAllocateTracks = 1,grid.col = tissue_colors_vibrant)

# Add labels for each tissue
circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
  tissue = CELL_META$sector.index
  circos.text(CELL_META$xcenter, CELL_META$ylim[1] + .1, tissue, 
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5))
}, bg.border = NA)


```

### Comparing the distributions
```{r,fig.dim=c(5,6)}
#Load all the data
tau_dat_all<-readRDS("./RDS/tau_dat_all.rds")
read_csv("./Datasets/DatasetS0_csv_files/Danio_biomart.csv")->Danio_biomart 
danio_conv<-read.csv("./Datasets/DatasetS0_csv_files/Danio_rerio_convergent_genes.csv")
read_tsv("./Datasets/DatasetS0_csv_files/phenoGeneCleanData_fish_2023.05.07.tsv")->pheno 
#Takes a long time
#pheno2 %>% mutate(Gene_symbol=str_extract(rem_str, paste(Danio_biomart$`Gene name`, collapse = "|")))->pheno2
#pheno2 %>% dplyr::rename("Gene Symbol"="Gene_symbol")->pheno2
#pheno2 %>% saveRDS("./RDS/pheno_with_clean_gene_symbol.rds")
readRDS("./RDS/pheno_with_clean_gene_symbol.rds")->pheno2

rbind(pheno %>% dplyr::select(c(`Gene Symbol`, `Affected Structure or Process 1 superterm ID`, `Publication ID`,`Affected Structure or Process 1 superterm Name`)),
      pheno2 %>%  dplyr::select(c(`Gene Symbol`, `Affected Structure or Process 1 superterm ID`, `Publication ID`,`Affected Structure or Process 1 superterm Name`))) %>% 
    group_by(`Gene Symbol`) %>% na.omit() %>% 
    mutate(uniq_affected_terms=length(unique(`Affected Structure or Process 1 superterm ID`)),
           uniq_pubs=length(unique(`Publication ID`))) %>% distinct(`Gene Symbol`, .keep_all = T)->pheno_summary

inner_join(
Danio_biomart %>% dplyr::select(`Gene stable ID`,`Gene name`,`Gene description`) %>% dplyr::rename("Gene.ID"=`Gene stable ID`,"Gene Symbol"=`Gene name`),
pheno_summary,by = "Gene Symbol") %>% na.omit() %>% distinct(Gene.ID,.keep_all = T)->pheno_summary

#inner_join(pheno_summary,
#estimated_ts_conv_Danio_tau_conv_all %>% dplyr::select(Gene.ID, conv), by = "Gene.ID")->pheno_summary

inner_join(pheno_summary,
           orth_list %>% dplyr::select(Orthogroup,orths) %>% dplyr::rename("Gene.ID"=orths),
           by="Gene.ID") %>% 
  dplyr::select(Gene.ID,Orthogroup,everything())->pheno_summary

#Summarise the average number of affected phenotypic terms by Orthogroup
#because each orthogroup has different genes with reporting of phenotypes
#Look at the above pheno_summary data and compare with the Danio_biomart df.

rbind(
  pheno_summary %>% summarise(n=mean(uniq_affected_terms),.by = c("Orthogroup")) %>% filter(Orthogroup %in% non_spuroius_conv$Orthogroup) %>% mutate(state="Convergent"),
  pheno_summary %>% summarise(n=mean(uniq_affected_terms),.by = c("Orthogroup")) %>% filter(!Orthogroup %in% non_spuroius_conv$Orthogroup) %>% mutate(state="Non-convergent"))->pheno_test_dat

q1<-quantile(log2(pheno_test_dat$n),.33)
q2<-quantile(log2(pheno_test_dat$n),.66)

pheno_test_dat %>% 
  ggplot(aes(x=log2(n), fill=state))+
  geom_density(alpha=0.6)+
  scale_fill_manual(values = c("#EA5B15","#1AAEC0"))+
  theme_bw()+
  geom_vline(xintercept = q1, linetype="dashed")+
  geom_vline(xintercept = q2, linetype="dashed")+
  xlab("Number of unique phenotypic terms (log2)")+
  labs(title = "Distribution of unique phenotypic terms\nduring genetic perpurbation.",
       subtitle = paste0("One-sided KS (greater) p-value: ",
       signif((ksboot(x=pheno_test_dat %>% filter(state=="Convergent") %>% pull(n),
        y=pheno_test_dat %>% filter(state=="Non-convergent") %>% pull(n),
        alternative = "greater"))$ksboot.pvalue, digits = 4)))->dist_mutant_phenotypes

pheno_test_dat %>% 
  ggplot(aes(x=log2(n), color = state))+
  stat_ecdf()+
  scale_color_manual(values = c("#EA5B15","#1AAEC0"))+
  theme_bw()+
  xlab("Number of unique phenotypic terms (log2)")+
  labs(title = "Cumulative probability distrubutions.")->ecdf_mutant_phenotypes

ggarrange(dist_mutant_phenotypes,ecdf_mutant_phenotypes, ncol = 1, common.legend = T)
```


## Pleiotropic release model {.tabset}
To reconcile our observation of multiple convergent substitutions in genes with intermediate levels of pleiotropy across distantly related fish species we introduce the pleiotropic release (PR) model (Fig 7A in main text). To test the PR model’s predictions, we use an evolutionary modeling approach based on Fisher’s geometric model (FGM). Using a distribution of pleiotropy derived from empirical data, we ran adaptive walks using two widely used evolutionary regimes: Strong Selection, Weak Mutation (SSWM) Regime, where mutations are rare but, when they occur, strong selection ensures the fixation of beneficial mutations and the loss of deleterious mutations. And the Greedy Adaptive Walk Regime, where multiple mutations arise (e.g., due to an increase in population size) and the mutation with the highest fitness advantage fixes in the population. The results of the adaptive walks showed that in the SSWM regime, the pleiotropy distribution of fixed mutations closely mirrored that of all mutations that arose, with slightly lower mean pleiotropy among fixed mutations, as expected due to limited mutation availability. In contrast, under the greedy walk regime, the pleiotropy distribution of fixed mutations shifted, favoring mutations with intermediate pleiotropy (Fig 7B). This suggests that in a mutation-rich environment, genes with intermediate pleiotropy play a crucial role in early adaptation to ecological changes. Code used to run the model can be found in the Scripts folder in the online repository in Zenodo.

### Load the data
```{r}
#File paths
files<-list.files(path = paste0("./Pleiotropy_data"), pattern = "\\.csv$", full.names = TRUE, recursive = TRUE)

#Sim types
scaled<-files[str_detect(files,"scaled")]
not_scaled<-files[str_detect(files,"scaled",negate = T)]

##Not scaled
sswm.not_scaled<-not_scaled[str_detect(not_scaled,"SSWM")]
greedy.not_scaled<-not_scaled[str_detect(not_scaled,"greedy")]

sim_features_greedy_sswm_not_scaled<-function(file_name){
  #Sim features for greedy and SSWM
  regime<-str_split(file_name,"_",simplify = T)[,2] %>% str_remove("data/")
  type_of_mut<-str_split(file_name,"_",simplify = T)[,3]
  dist<-str_split(file_name,"_",simplify = T)[,5] %>% str_remove("dist=")
  ntraits<-str_split(file_name,"_",simplify = T)[,6] %>% str_remove("ntraits=")
  
  read.csv(file_name, col.names = "P") %>% mutate(regime=regime, type_of_mut=type_of_mut, dist=dist, ntraits,ntraits) %>% 
    as.data.frame() %>% return()
}


purrr::map(sswm.not_scaled,
           sim_features_greedy_sswm_not_scaled) %>% list_rbind()->sswm.not_scaled_out


purrr::map(greedy.not_scaled,
           sim_features_greedy_sswm_not_scaled) %>% list_rbind()->greedy.not_scaled_out

```

### L distribution
```{r}
##Unscaled plots
sswm.not_scaled_out %>% filter(dist=="L") %>% filter(ntraits==10) %>% 
  ggplot(aes(x=P, fill = type_of_mut, y=after_stat(density)))+
  facet_grid(rows=vars(type_of_mut))+
  geom_histogram(color="black", binwidth = 1)+
  scale_fill_manual(values = c("#4A72EA","#EAC34A"))+
  labs(title = "SSWM regime")+
  xlab("Pleiotropy of mutation")+
  theme_bw()+
  theme(legend.position = "none",
        plot.title = element_text(size = 8),
        axis.title = element_text(size = 8))->sswm_not_scaled_L

greedy.not_scaled_out %>% filter(dist=="L") %>% filter(ntraits==10) %>% 
  ggplot(aes(x=P, fill = type_of_mut, y=after_stat(density)))+
  facet_grid(rows=vars(type_of_mut))+
  geom_histogram(color="black", binwidth = 1)+
  scale_fill_manual(values = c("#4A72EA","#EAC34A"))+
  labs(title = "Greedy walk regime")+
  xlab("Pleiotropy of mutation")+
  theme_bw()+ 
  theme(legend.position = "none",
        plot.title = element_text(size = 8),
        axis.title = element_text(size = 8))->greedy_not_scaled_L


ggpubr::ggarrange(sswm_not_scaled_L, greedy_not_scaled_L, ncol = 2)

kldtools::ksboot(x = greedy.not_scaled_out %>% filter(dist=="L") %>% filter(ntraits==10) %>% filter(type_of_mut=="all") %>% pull(P),
        y = greedy.not_scaled_out %>% filter(dist=="L") %>% filter(ntraits==10) %>% filter(type_of_mut=="fixed") %>% pull(P), alternative = "greater",nboots = 1000)
```

### Reverse L
```{r}
##Other distributions
#Reverse L
sswm.not_scaled_out %>% filter(dist=="rev.L") %>% filter(ntraits==10) %>% 
  ggplot(aes(x=P, fill = type_of_mut, y=..density..))+
  facet_grid(rows=vars(type_of_mut))+
  geom_histogram(color="black", binwidth = 1)+
  scale_fill_manual(values = c("#4A72EA","#EAC34A"))+
  labs(title = "SSWM regime")+
  xlab("Pleiotropy of mutation")+
  theme_bw()+
  theme(legend.position = "none",
        plot.title = element_text(size = 8),
        axis.title = element_text(size = 8))->sswm_not_scaled_rev.L

greedy.not_scaled_out %>% filter(dist=="rev.L") %>% filter(ntraits==10) %>% 
  ggplot(aes(x=P, fill = type_of_mut, y=..density..))+
  facet_grid(rows=vars(type_of_mut))+
  geom_histogram(color="black", binwidth = 1)+
  scale_fill_manual(values = c("#4A72EA","#EAC34A"))+
  labs(title = "Greedy walk regime")+
  xlab("Pleiotropy of mutation")+
  theme_bw()+ 
  theme(legend.position = "none",
        plot.title = element_text(size = 8),
        axis.title = element_text(size = 8))->greedy_not_scaled_rev.L

ggpubr::ggarrange(sswm_not_scaled_rev.L, greedy_not_scaled_rev.L, ncol = 2)
```

### Uniform
```{r}

sswm.not_scaled_out %>% filter(dist=="uni") %>% filter(ntraits==10) %>% 
  ggplot(aes(x=P, fill = type_of_mut, y=..density..))+
  facet_grid(rows=vars(type_of_mut))+
  geom_histogram(color="black", binwidth = 1)+
  scale_fill_manual(values = c("#4A72EA","#EAC34A"))+
  labs(title = "SSWM regime")+
  xlab("Pleiotropy of mutation")+
  theme_bw()+
  theme(legend.position = "none",
        plot.title = element_text(size = 8),
        axis.title = element_text(size = 8))->sswm_not_scaled_uni

greedy.not_scaled_out %>% filter(dist=="uni") %>% filter(ntraits==10) %>% 
  ggplot(aes(x=P, fill = type_of_mut, y=..density..))+
  facet_grid(rows=vars(type_of_mut))+
  geom_histogram(color="black", binwidth = 1)+
  scale_fill_manual(values = c("#4A72EA","#EAC34A"))+
  labs(title = "Greedy walk regime")+
  xlab("Pleiotropy of mutation")+
  theme_bw()+ 
  theme(legend.position = "none",
        plot.title = element_text(size = 8),
        axis.title = element_text(size = 8))->greedy_not_scaled_uni

ggpubr::ggarrange(sswm_not_scaled_uni, greedy_not_scaled_uni, ncol = 2)
```

### All distribution plots
```{r,fig.dim=c(5,10)}
  ggpubr::ggarrange(ggpubr::ggarrange(sswm_not_scaled_L, greedy_not_scaled_L, ncol = 2),  
                    ggpubr::ggarrange(sswm_not_scaled_rev.L, greedy_not_scaled_rev.L, ncol = 2),  
                    ggpubr::ggarrange(sswm_not_scaled_uni, greedy_not_scaled_uni, ncol = 2), 
                    ncol = 1)

```



