---
title: "Estimates of molecular convergence reveal pleiotropic genes underlying adaptive variation across teleost fish - Report"
author: "Agneesh Barua"
date: "17/03/2025"
output: 
  html_document:
    code_folding: hide
    theme: 'lumen'
    df_print: 'paged'
    toc: TRUE
    toc_float: TRUE
---

This report comprises R code used in the current manuscript. Users can load the appropriate Rmd files found in the electronic supplementary material (DatasetS10_Scripts) and reproduce the analysis and steps showed here. 
Electronic supplementary material containing all the data files can be found at: https://zenodo.org/records/12645625
  
Users are free to use any piece of code in heir own studies. If you use any of the code here please cite:  
Estimates of molecular convergence reveal pleiotropic genes underlying adaptive variation across teleost fish. 
doi: https://doi.org/10.1101/2024.06.24.600426

```{r child = '00.Orthologs_estimates.Rmd'}
```

```{r child = '01.Getting_bgee_dat.Rmd'}
```

```{r child = '02.csubst_analysis.Rmd'}
```

```{r child = '03.Gene_expression_analysis.Rmd'}
```

```{r child = '04.Revised_analysis.Rmd'}
```
